Optimizing Abstract Abstract Machines

### 2013-06-17 paper/paper.tex
More reviewer feedback incorporated. Need to start trimming back to 12 pages.

%\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
\title{Correctly Optimizing Abstract Abstract Machines}

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator\footnote{Technically, we would have to copy the value of the global store to all states being stepped to fit the formal definition of a widening, but the consequential accelerated convergence fits the M.O. of a widening operator.} during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, complexity drops from intractable exponentiality to a merely
daunting polynomial.

@@

The final approximation we make to get to our baseline is to
store-allocate all values that appear, so that any non-machine state
that contains a value instead contains an address to a value.  The AAM
approach stops at the previous optimization.  However, the {\tt fn}
continuation stores a value, and this makes the space of continuations
quadratic rather than linear in the size of the program, for a
monovariant analysis like 0-CFA.  Having the space of continuations
grow linearly with the size of the program will drop the overall
complexity to cubic (as expected).

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially (${\mathcal O}(n^n)$). We extend the semantics to
additionally allocate an address for the function value when creating
the $\kfn{}$ continuation. The continuation has to contain this address
to remember where to retrieve values from in the store.

The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.
\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
\text{ where }
  \maddr &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision:
\begin{align*}
\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr)
\end{align*}
Associated with this indirection, we now apply all functions stored in
the address. This nondeterminism is necessary in order to continue with
evaluation.

@@

The state space changes from a store and set of contexts to a set
of seen abstract states (context plus store), $S$, a set of contexts to
step (the frontier), $F$, and a store to step those contexts with, $\msto$:
\begin{equation*}
(S, F, \msto) \mathrel{\widehat{\machstep}} (S \cup S', F', \msto')
\end{equation*}

We constantly see more states, so $S$ is always growing. The frontier,
which is what remains to be done, changes. Let's start with the result
of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):
%
\begin{equation*}
I = \setof{(c',\msto') \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto'), c \in F}
\end{equation*}

@@

\begin{equation*}
\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
\end{equation*}
%
The next frontier is exactly the states that we found from stepping
the last frontier, but have not seen before. They must be
\emph{states}, so we pair the contexts with the next store:
%
\begin{equation*}
F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin S}
\end{equation*}
%
Finally, we add what we know we had not yet seen to the seen set:
\begin{equation*}
  S' = \setof{(c,\msto') \mid c \in F'}
\end{equation*}
%
To inject a program $e$ into this machine, we start off knowing we
have seen the first state, and that we need to process the first
state:
\begin{align*}
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

Notice that now $S$ has several copies of the abstract store in it. As
it is, this semantics is much less efficient (but still more precise)
than the previously proposed semantics because membership checks have
to compare entire stores. Checking equality is expensive because the stores within each are
large, and every entry must be checked against every other. Hashes can
sometimes rule out inequality relatively quickly, but the incidence of
collisions and actual equality is costly.

@@

\begin{enumerate}
\item{In our semantics, the store does not change until the entire
    frontier has been explored. This avoids cross-branch pollution
    which would otherwise happen in Shivers' semantics, e.g., when one branch writes to
    address $\maddr$ and another branch reads from address
    $\maddr$.}
\item{The common implementation strategy for timestamps destructively
    updates each state's timestamp. This loses \emph{temporal}
    information about the contexts a state is visited in, and in what
    order. Our semantics has a drop-in replacement of timestamps for
    stores in the seen set ($S$), so we do not experience precision
    loss.}
\end{enumerate}

{\small
\begin{align*}
\Sigma \in \Store^* \qquad S \subseteq {\mathbb N} \times \Context \qquad F \subseteq \Context
\end{align*}
\begin{align*}
(S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
\mbox{where } I &= \setof{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
              (t',\Sigma') &=\left\{\begin{array}{ll}
                           (t+1,\msto'\Sigma') & \text{ if } \msto' \neq \msto \\
                           (t,\Sigma)   & \text{ otherwise}
                          \end{array}\right. \\
              F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
              S' &= \setof{(c,t') \mid c \in F'}\\
\inject(e) &= (\setof{\ttuple{c_0}{0}},\setof{c_0},\bot,\cons{\bot}{\epsilon},0) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}}

@@

\begin{lemma}
  $\widehat{\machstep}$ maintains the invariant that all stores in $S$ are
  totally ordered and $\msto$ is an upper bound of the stores in $S$.
\end{lemma}

@@

\begin{theorem}
$\widehat{\machstep}^T$ is a complete abstraction of $\widehat{\machstep}$.
\end{theorem}
The proof follows from the order isomorphism that, in one direction,
sorts all the stores in $S$ to form $\Sigma$, and translates stores in
$S$ to their distance from the end of $\Sigma$ (their timestamp). In the other
direction, timestamps in $S$ are replaced by the stores they point to in $\Sigma$.

@@

We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
 becomes a log addition
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
empty ($\mtlst$) for each step. Applying the changes to the full store
is straightforward:
\begin{equation*}
\replay : (\StoreDelta \times \Store) \to (\Store \times \Boolean)
\end{equation*}
\begin{align*}
\replay(\left[ \ttuple{\maddr_i}{\mval{s_i}}, \ldots\right], \msto) &=
\ttuple{\msto'}
       {\mval{s_i} \overset{?}{=} \msto'(\maddr_i) \vee \ldots} \\
\text{ where } \msto' &= {\msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots}
\end{align*}

@@

A taste of the changes to the reduction relation is as follows:
\begin{equation*}
\dmachstep \subseteq (\Context\times\Store) \times (\Context\times\StoreDelta) 
\end{equation*}
\begin{align*}
\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
\text{ where }\maddr &= \alloc(\mstate) \\
              \menv' &= \menv[\mvar\mapsto\maddr]
\end{align*}

@@

We lift $\dmachstep$ to accommodate for the asymmetry
in the input and output, and change the frontier-based semantics in
the following way:
{\small
\begin{align*}
(S, F, \msto,\Sigma,t) &\mathrel{\damachstep} (S \cup S', F', \msto',\Sigma',t') \\
\mbox{ where }
 I &= \setof{(c',\msdiff) \mid \ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff} } \\
 \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 \ttuple{t'}{\Sigma'} &=
     \left\{
       \begin{array}{ll}
         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
         \ttuple{t}{\Sigma} & \text{ otherwise}
       \end{array}\right. \\
 F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
 S' &= \setof{(c,t') \mid c \in F'} \\
\appendall(\varnothing) &= \epsilon \\
\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
\end{align*}}

@@

Tracing the execution of the analysis reveals an immediate shortcoming:
there is a high degree of branching and merging in the exploration.
%
Surveying this branching has no benefit for precision.
%
For example, in a function application, {\tt (f x y)},
where {\tt f}, {\tt x} and {\tt y} each have several values
each argument evaluation induces $n$-way branching, only to be ultimately joined back together in their respective
application positions.
%
Transition patterns of this shape litter the state-graph:
%
\begin{center}
\includegraphics[scale=0.3]{fanout}
\end{center}
To avoid the spurious forking and joining, we {\it delay} the non-determinism
until and unless it is needed in {\it strict contexts} (such as the guard for an
{\tt if}, a called procedure, or a numerical primitive application).
%
Doing so collapses these forks and joins into a linear sequence of states:
\begin{center}
\includegraphics[scale=0.3]{lazy}
\end{center}

@@

\paragraph{Option 1: Lose precision; simplify implementation}
This semantics introduces a subtle precision difference over the
baseline. Consider a configuration where a reference to a variable and
a binding of a variable will happen in one step, since store widening
leads to stepping several states in one big ``step.'' With laziness,
the reference will mean the original binding(s) of the variable or the
new binding, because the actual store lookup is delayed one step
(i.e. laziness is administrative). Without laziness, the reference
will fan out to all the bindings of the variable before the new
binding happens and thus theoretically has an observable precision
difference. % All our benchmarks maintain their precision with lazy non-determinism, however.

\paragraph{Option 2: Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by storing the result of the lookup in a value
representing a delayed nondeterministic choice. This has problems with
the next optimization, leading to unsoundness. Regaining soundness
would involve artificial complications to the semantics.

\paragraph{Our choice: option 1}
The configurations that lead to precision loss happen too rarely to
warrant the significant increase in time and memory needed for this
eager non-determinism. Without a widened store, lazy
non-determinism is not a complete abstraction with either approach,
but the precision gains are not immediately apparent due to fake
rebinding (a second reference to the same variable chooses a different
value from the first) and other problems. The performance and memory
gains are apparent, so we favor the simplicity over invisible
gains.}{}

@@

The compilation step converts expressions into functions that expect
the other components of the {\tt ev} state. Its definition in figure
\ref{fig:compile} shows close similarity to the rules for interpreting
{\tt ev} states. The next step is to change reduction rules that
create {\tt ev} states to instead call these functions. Figure
\ref{fig:caam} shows the modified reduction relation. The only change
from the previous semantics is that $\ev{}$ state construction is
replaced by calling the compiled expression. For notational coherence,
we write $\lambda^\mcntr(\mathit{args} \ldots)$ for
$\lambda(\mathit{args} \ldots, \mcntr)$ and
$\mcomp^\mcntr(\mathit{args}\ldots)$ for $\mcomp(\mathit{args}\ldots,
\mcntr)$.

@@

\begin{figure}
\small
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \setof{ \inject(\compile{\mexp}^\mtcntr_\bot(\bot,\epsilon,\kmt)) \multimachstep \mstate}
                    \text{ where } \\
\inject(c,\msdiff) &= \wn(c,\replay(\msdiff,\bot)) \\
\wn(c,\msto) \machstep \wn(c',\msto') &\iff c \cmachstep_\msto c',\msdiff \\
\msdiff \text{ is such that } &\replay(\msdiff,\msto) = \msto'
\end{align*}
\\[2mm]
\begin{align*}
%% CONTINUE
%%\co{\kmt,\mval} &\cmachstep_\msto \ans{\mval'},\epsilon \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
%%\\
\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\mkaddr},\mval} & \cmachstep_\msto
\mcomp^\mcntr(\msto)(\menv,\msdiff,\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}) \\
\text{ where } \maddr_f &= \alloc(\mstate) \\
               \msdiff &= \cons{\ttuple{\maddr_f}{\force(\msto,\mval)}}{\epsilon}
\\
\co{\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr},\mval} & \cmachstep_\msto
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont},\epsilon \\
\text{ if }\mvalx{u} &\in \msto(\maddr_f), \mkont \in \msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\mkaddr},\strue} & \cmachstep_\msto
\mcompi{0}^\mcntr(\msto)(\menv,\epsilon,\mkont)
\text{ if }\mkont\in\msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\mkaddr},\sfalse} & \cmachstep_\msto
\mcompi{1}^\mcntr(\msto)(\menv,\epsilon,\mkont)
\text{ if }\mkont\in\msto(\mkaddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\mkont} & \cmachstep_\msto
\mcomp^{\mcntr'}(\msto)(\menv',\msdiff,\mkont) \\
\text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
              \msdiff &= \cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\epsilon}
\\
\ap{\mop,\mval,\mkont} & \cmachstep_\msto
\co{\mkont,\mval'},\epsilon \\
\text{ where } \mvalx{u} &\in \spchoice{\force(\msto,\mval)}{\force(\mval)}, \mval'\in\interpdelta(\mop,\mvalx{u})
\end{align*}
\end{gather*}
\caption{Abstract abstract machine for compiled ISWIM}
\label{fig:caam}
\end{figure}

@@

The number of transitions that can occur in succession from an
abstract bytecode is roughly bounded by the amount of expression
nesting in the program. The partial order that defines subexpressions
smaller than the expression in which they occur is well-founded, so we
use this observation to drive our use of a well-founded equivalence
bisimulation (WEB)~\cite{ianjohnson:manolios-diss}. WEBs are equivalent to the notion of a stuttering
bisimulation, but are more amenable to mechanisation (and thus
rigorous proof). They also only require reasoning over one step of the
reduction relation, which makes the proof concise.

We define a refinement from non-compiled to compiled states by
``committing'' all the actions of an $\ev{}$ state (defined similarly to
$\compile{\_}$, but immediately applies the functions), and
subsequently changing all expressions with their compiled
variants. Since WEBs are for single transition systems, a WEB
refinement is over the disjoint union of our two semantics, and the
equivalence relation we use is just that a state is related to its
refined state (and itself). Call this relation $B$.

Before we prove this setup is indeed a WEB, we need one lemma that
applying an abstract bytecode's function is equal to refining the
corresponding $\ev{}$ state:
%
\begin{lemma}[Compile/commit]
Let $c,\msdiff' = \compile{\mexp}^\mcntr_{r(\msto)}(\menv,\msdiff,r(\mkont))$.
Let $\wn(c',\msto') = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
$\wn(c,\replay(\msdiff',\msto)) = \wn(c',\replay(\msdiff,\msto'))$.
\end{lemma}
The proof is by induction on $\mexp$.

@@

Thus far, we have made our optimizations in a purely functional
manner. For the final push for performance, we need to dip into the
imperative. In this section, we show an alternative representation of
the store and seen set that are more space-efficient and are amenable
to destructive updates.

@@

The store maps to a stack of timestamped sets of abstract values.

@@

Here we formally define what lookup and update mean at a given
timestamp.

>>>>

%\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
\title{Optimizing Abstract Abstract Machines}

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as
the application of a widening operator\footnote{Technically, we would
  have to copy the value of the global store to all states being
  stepped to fit the formal definition of a widening, but this
  representation is order-isomorphic to that.} during the fixed-point
iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, complexity drops from intractable exponentiality to a merely
daunting polynomial.

@@

The final approximation we make to get to our baseline is to
store-allocate all values that appear, so that any non-machine state
that contains a value instead contains an address to a value.  The AAM
approach stops at the previous optimization.  However, the \kfnalone
continuation stores a value, and this makes the space of continuations
quadratic rather than linear in the size of the program, for a
monovariant analysis like 0-CFA.  Having the space of continuations
grow linearly with the size of the program will drop the overall
complexity to cubic (as expected). We also need to allocate an address
for the argument position in an \apalone state.

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially (${\mathcal O}(n^n)$). We extend the semantics to
additionally allocate an address for the function value when creating
the $\kfnalone$ continuation. The continuation has to contain this address
to remember where to retrieve values from in the store.

The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.
\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
\text{ where }
  \maddr &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision:
\begin{align*}
\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\maddr,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr) \\
\text{ where } \maddr &= \alloc(\mstate) \\
               \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}

Associated with this indirection, we now apply all functions stored in
the address. This nondeterminism is necessary in order to continue with
evaluation.

@@

The state space changes from a store and set of contexts to a set
of seen abstract states (context plus store), $\mseen$, a set of contexts to
step (the frontier), $F$, and a store to step those contexts with, $\msto$:
\begin{equation*}
(\mseen, F, \msto) \mathrel{\widehat{\machstep}} (\mseen \cup \mseen', F', \msto')
\end{equation*}

We constantly see more states, so $\mseen$ is always growing. The frontier,
which is what remains to be done, changes. Let's start with the result
of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):
%
\begin{equation*}
I = \setof{(c',\msto') \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto'), c \in F}
\end{equation*}

@@

\begin{equation*}
\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
\end{equation*}
%
The next frontier is exactly the states that we found from stepping
the last frontier, but have not seen before. They must be
\emph{states}, so we pair the contexts with the next store:
%
\begin{equation*}
F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin \mseen}
\end{equation*}
%
Finally, we add what we know we had not yet seen to the seen set:
\begin{equation*}
  \mseen' = \setof{(c,\msto') \mid c \in F'}
\end{equation*}
%
To inject a program $e$ into this machine, we start off knowing we
have seen the first state, and that we need to process the first
state:
\begin{align*}
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

Notice that now $\mseen$ has several copies of the abstract store in it. As
it is, this semantics is much less efficient (but still more precise)
than the previously proposed semantics because membership checks have
to compare entire stores. Checking equality is expensive because the stores within each are
large, and every entry must be checked against every other. Hashes can
sometimes rule out inequality relatively quickly, but the incidence of
collisions and actual equality is costly.

@@

\begin{enumerate}
\item{In our semantics, the store does not change until the entire
    frontier has been explored. This avoids cross-branch pollution
    which would otherwise happen in Shivers' semantics, e.g., when one branch writes to
    address $\maddr$ and another branch reads from address
    $\maddr$.}
\item{The common implementation strategy for timestamps destructively
    updates each state's timestamp. This loses \emph{temporal}
    information about the contexts a state is visited in, and in what
    order. Our semantics has a drop-in replacement of timestamps for
    stores in the seen set ($\mseentime$), so we do not experience precision
    loss.}
\end{enumerate}

{\small
\begin{align*}
\Sigma \in \Store^* \qquad \mseentime \subseteq {\mathbb N} \times \Context \qquad F \subseteq \Context
\end{align*}
\begin{align*}
(\mseentime, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (\mseentime \cup \mseentime', F', \msto', \Sigma', t') \\
\mbox{where } I &= \setof{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
              (t',\Sigma') &=\left\{\begin{array}{ll}
                           (t+1,\msto'\Sigma') & \text{ if } \msto' \neq \msto \\
                           (t,\Sigma)   & \text{ otherwise}
                          \end{array}\right. \\
              F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin \mseentime} \\
              \mseentime' &= \setof{(c,t') \mid c \in F'}\\
\inject(e) &= (\setof{\ttuple{c_0}{0}},\setof{c_0},\bot,\cons{\bot}{\epsilon},0) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}}

@@

\begin{lemma}
  $\widehat{\machstep}$ maintains the invariant that all stores in $\mseen$ are
  totally ordered and $\msto$ is an upper bound of the stores in $\mseen$.
\end{lemma}

@@

\begin{theorem}
$\widehat{\machstep}^T$ is a complete abstraction of $\widehat{\machstep}$.
\end{theorem}
The proof follows from the order isomorphism that, in one direction,
sorts all the stores in $\mseen$ to form $\Sigma$, and translates stores in
$\mseen$ to their distance from the end of $\Sigma$ (their timestamp). In the other
direction, timestamps in $\mseentime$ are replaced by the stores they point to in $\Sigma$.

@@

We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
 becomes a log addition
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
empty ($\mtlst$) for each step. Applying the changes to the full store
is straightforward:
\begin{equation*}
\replay : (\StoreDelta \times \Store) \to (\Store \times \Boolean)
\end{equation*}
\begin{align*}
\replay(\left[ \ttuple{\maddr_i}{\mval{s_i}}, \ldots\right], \msto) &=
\ttuple{\msto'}
       {\diffp(\mval{s_i}, \msto(\maddr_i)) \vee \ldots} \\
\text{ where } \msto' &= {\msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots}
\diffp(\mval{s}, \mval{s'}) &= \mval{s'} \overset{?}{=} \mval{s} \sqcup \mval{s'}
\end{align*}

@@

A taste of the changes to the reduction relation is as follows:
\begin{equation*}
\dmachstep \subseteq (\Context\times\Store) \times (\Context\times\StoreDelta) 
\end{equation*}
\begin{align*}
\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\maddr,\mkont}}{\msto} & \dmachstep
\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr'}{\msto(\maddr)}}{\epsilon}} \\
\text{ where }\maddr' &= \alloc(\mstate) \\
              \menv' &= \menv[\mvar\mapsto\maddr']
\end{align*}

@@

We lift $\dmachstep$ to accommodate for the asymmetry
in the input and output, and change the frontier-based semantics in
the following way:
{\small
\begin{align*}
(\mseentime, F, \msto,\Sigma,t) &\mathrel{\damachstep} (\mseentime \cup \mseentime', F', \msto',\Sigma',t') \\
\mbox{ where }
 I &= \setof{(c',\msdiff) \mid \ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff} } \\
 \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 \ttuple{t'}{\Sigma'} &=
     \left\{
       \begin{array}{ll}
         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
         \ttuple{t}{\Sigma} & \text{ otherwise}
       \end{array}\right. \\
 F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin \mseentime} \\
 \mseentime' &= \setof{(c,t') \mid c \in F'} \\
\appendall(\varnothing) &= \epsilon \\
\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
\end{align*}}

@@

Tracing the execution of the analysis reveals an immediate shortcoming:
there is a high degree of branching and merging in the exploration.
%
Surveying this branching has no benefit for precision.
%
For example, in a function application, {\tt (f x y)},
where {\tt f}, {\tt x} and {\tt y} each have several values
each argument evaluation induces $n$-way branching, only to be ultimately joined back together in their respective
application positions.
%
Transition patterns of this shape litter the state-graph:
%
\begin{center}
\includegraphics[scale=0.2]{fanoutdot2}
\end{center}
To avoid the spurious forking and joining, we {\it delay} the non-determinism
until and unless it is needed in {\it strict contexts} (such as the guard for an
{\tt if}, a called procedure, or a numerical primitive application).
%
Doing so collapses these forks and joins into a linear sequence of states:
\begin{center}
\includegraphics[scale=0.2]{lazydot}
\end{center}

@@

\paragraph{Option 1: Lose precision; simplify implementation}
This semantics introduces a subtle precision difference over the
baseline. Consider a configuration where a reference to a variable and
a binding of a variable will happen in one step, since store widening
leads to stepping several states in one big ``step.'' With laziness,
the reference will mean the original binding(s) of the variable or the
new binding, because the actual store lookup is delayed one step
(i.e. laziness is administrative).

\paragraph{Option 2: Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by storing the result of the lookup in a value
representing a delayed nondeterministic choice. This is a more common
choice in 0CFA implementations we have seen, but it interferes with
the next optimization (discussion in next section).

\paragraph{Our choice: option 1}
The conditions for precision loss happen too rarely to warrant the
added complexity to the next section's semantics to avoid such
losses. Without abstract compilation, option 2 is the better
choice.}{}

@@

The compilation step converts expressions into functions that expect
the other components of the {\tt ev} state. Its definition in figure
\ref{fig:compile} shows close similarity to the rules for interpreting
{\tt ev} states. The next step is to change reduction rules that
create {\tt ev} states to instead call these functions. Figure
\ref{fig:caam} shows the modified reduction relation. The only change
from the previous semantics is that $\evalone$ state construction is
replaced by calling the compiled expression. For notational coherence,
we write $\lambda^\mcntr(\mathit{args} \ldots)$ for
$\lambda(\mathit{args} \ldots, \mcntr)$ and
$\mcomp^\mcntr(\mathit{args}\ldots)$ for $\mcomp(\mathit{args}\ldots,
\mcntr)$.

@@

\begin{figure}
\small
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \setof{ \inject(\compile{\mexp}^\mtcntr_\bot(\bot,\epsilon,\kmt)) \multimachstep \mstate}
                    \text{ where } \\
\inject(c,\msdiff) &= \wn(c,\replay(\msdiff,\bot)) \\
\wn(c,\msto) \machstep \wn(c',\msto') &\iff c \cmachstep_\msto c',\msdiff \\
\msdiff \text{ is such that } &\replay(\msdiff,\msto) = \msto'
\end{align*}
\\[2mm]
\begin{align*}
%% CONTINUE
%%\co{\kmt,\mval} &\cmachstep_\msto \ans{\mval'},\epsilon \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
%%\\
\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\mkaddr},\mval} & \cmachstep_\msto
\mcomp^\mcntr(\msto)(\menv,\msdiff,\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}) \\
\text{ where } \maddr_f &= \alloc(\mstate) \\
               \msdiff &= \cons{\ttuple{\maddr_f}{\force(\msto,\mval)}}{\epsilon}
\\
\co{\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr},\mval} & \cmachstep_\msto
\ap[^\mcntr_\mlab]{\mvalx{u},\maddr,\mkont},\cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\epsilon} \\
\text{ if }\mvalx{u} &\in \msto(\maddr_f), \mkont \in \msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\mkaddr},\strue} & \cmachstep_\msto
\mcompi{0}^\mcntr(\msto)(\menv,\epsilon,\mkont)
\text{ if }\mkont\in\msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\mkaddr},\sfalse} & \cmachstep_\msto
\mcompi{1}^\mcntr(\msto)(\menv,\epsilon,\mkont)
\text{ if }\mkont\in\msto(\mkaddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\maddr,\mkont} & \cmachstep_\msto
\mcomp^{\mcntr'}(\msto)(\menv',\msdiff,\mkont) \\
\text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
              \msdiff &= \cons{\ttuple{\maddr}{\msto(\maddr)}}{\epsilon}
\\
\ap{\mop,\maddr,\mkont} & \cmachstep_\msto
\co{\mkont,\mvalx{u}},\epsilon \\
\text{ where } \mval &\in \spchoice{\msto(\maddr)}{\msto(\maddr)}, \mvalx{u}\in\interpdelta(\mop,\mval)
\end{align*}
\end{gather*}
\caption{Abstract abstract machine for compiled ISWIM}
\label{fig:caam}
\end{figure}

@@

The number of transitions that can occur in succession from an
abstract bytecode is roughly bounded by the amount of expression
nesting in the program. The partial order that defines subexpressions
smaller than the expression in which they occur is well-founded, so we
use this observation to drive our use of a well-founded equivalence
bisimulation (WEB)~\cite{ianjohnson:manolios-diss}. WEBs are
equivalent to the notion of a stuttering bisimulation, but are more
amenable to mechanisation (and thus rigorous proof). They also only
require reasoning over one step of the reduction relation, which makes
the proof concise. The trick is in defining a well-founded ordering
that determines when the two semantics will match up again, what
Manolios calls the pair of functions $\erankt$ and $\erankl$ (but we
don't need the generality that requires $\erankl$ since the uncompiled
semantics doesn't stutter at all).

We define a refinement, $r$, from non-compiled to compiled states (built structurally) by
``committing'' all the actions of an $\evalone$ state (defined similarly to
$\compile{\_}$, but immediately applies the functions), and
subsequently changing all expressions with their compiled
variants. Since WEBs are for single transition systems, a WEB
refinement is over the disjoint union of our two semantics, and the
equivalence relation we use is just that a state is related to its
refined state (and itself). Call this relation $B$.

Before we prove this setup is indeed a WEB, we need one lemma that
applying an abstract bytecode's function is equal to refining the
corresponding $\evalone$ state:
%
\begin{lemma}[Compile/commit]
Let $c,\msdiff' = \compile{\mexp}^\mcntr_{r(\msto)}(\menv,\msdiff,r(\mkont))$.
Let $\wn(c',\msto') = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
$\wn(c,\replay(\msdiff',\msto)) = \wn(c',\replay(\msdiff,\msto'))$.
\end{lemma}
The proof is by induction on $\mexp$.

@@

Thus far, we have made our optimizations in a purely functional
manner. For the final push for performance, we need to dip into the
imperative. In this section, we show an alternative representation of
the store and seen set that are more space-efficient and are amenable
to destructive updates by adhering to a history for each address.

@@

The store maps to a stack of timestamped sets of abstract
values. Throughout this section, we will be taking the parameter $t$
to be the ``current time,'' or the length of the store chain at the
beginning of the step.

@@

To allow imperative store updates, we maintain an invariant that we never look up values tagged at a time in the future:
